Morphological diversity in tenrecs (Afrosoricida, Tenrecidae): Comparing tenrec skull diversity to their closest relatives

### 2015-03-12 paper/Manuscript.tex
Changes following reviewers' comments

	Unfortunately, morphological diversity is difficult to quantify. Many studies are constrained to measuring the diversity of specific traits rather than overall morphologies \citep{Roy1997}. In addition, our perception of morphological diversity is influenced by the trait being used, and results may depend on the particular trait being analysed \citep{Foth2012}. Furthermore, linear measurements of morphological traits can restrict our understanding of overall morphological variation; a distance matrix of measurements among specific points is unlikely to give a completely accurate representation of a three dimensional structure \citep{Rohlf1993}. Geometric morphometric approaches can circumvent some of these issues by using a system of Cartesian landmark coordinates to define anatomical points \citep{Adams2004}. This method captures more of the true, overall anatomical shape of particular structures \citep{Mitteroecker2009}. In particular, two-dimensional geometric morphometric approaches are commonly used to analyse 3D morphological shape and are appropriate for cross-species comparisons \citep[e.g.][]{Muschick2012, Panchetti2008, Wroe2007}. Any bias from 2D representation of a 3D structure is unlikely to be a significant issue for interspecific studies as the overall shape variation among species is geater than discrepancies introduced by using 2D morphometric techniques \citep{Cardinin2014}. These more detailed approaches are useful tools for studying patterns of morphological diversity.

	Here we apply geometric morphometric techniques to quantify morphological diversity in a family of small mammals, the tenrecs. Tenrecs (Afrosoricida, Tenrecidae) are a morphologically diverse group that researchers often identify as an example of both convergent evolution and an adaptive radiation \citep{Soarimalala2011, Eisenberg1969}. The family is comprised of 34 species, 31 of which are endemic to Madagascar \citep{Olson2013}. Body masses of tenrecs span three orders of magnitude (2.5 to $>$ 2,000g); a greater range than all other Families, and most Orders, of living mammals \citep{Olson2003}. Within this vast size range there are tenrecs which convergently resemble shrews (\textit{Microgale} tenrecs), moles (\textit{Oryzorictes} tenrecs) and hedgehogs \citep[\textit{Echinops} and \textit{Setifer} tenrecs,][]{Eisenberg1969}. Their similarities include examples of morphological, behavioural and ecological convergence \citep{Soarimalala2011}. Tenrecs are one of only four endemic mammalian clades in Madagascar and the small mammal species they resemble are absent from the island \citep{Garbutt1999}. Therefore, it appears that tenrecs represent an adaptive radiation of species which filled otherwise vacant ecological niches \citep{Soarimalala2011}.

	Although tenrecs are often cited as an example of both an adaptive radiation and exceptional convergent evolution \citep{Soarimalala2011, Eisenberg1969}, these claims have not been investigated quantitatively. There are qualitative similarities among the hind limb morphologies of tenrecs and several other unrelated species with similar locomotory styles \citep{Salton2009} but the degree of morphological similarity has not been established. Morphological diversity is an important feature of adaptive radiations \citep{Losos2010a} and it also informs our understanding of convergent phenotypes \citep{Muschick2012}. Therefore, it is important to quantify patterns of morphological diversity in tenrecs to gain an insight into their evolution.

	We present the first quantitative study of patterns of morphological diversity in tenrecs. We use geometric morphometric techniques \citep{Rohlf1993} to compare cranial morphological diversity in tenrecs to that of their closest relatives, the golden moles (Afrosoricida, Chrysochloridae). 
	We expect tenrecs to be more morphologically diverse than golden moles because tenrecs occupy a wider variety of ecological niches. The tenrec family includes terrestrial, semi-fossorial, semi-aquatic and semi-arboreal species \citep{Soarimalala2011}. In contrast, all golden moles occupy very similar, fossorial ecological niches \citep{Bronner1995}. Greater ecological variety is often (though not always) correlated with higher morphological diversity \citep{Losos2010a}. However, our results reveal that, in skulls at least, morphological diversity in tenrecs is not as great as it first appears.

@@

\subsection{Data collection}
	One of us (SF) collected data from five museums: Natural History Museum, London (BMNH), Smithsonian Institute Natural History Museum, Washington D.C. (SI), American Museum of Natural History, New York (AMNH), Museum of Comparative Zoology, Cambridge M.A. (MCZ) and Field Museum of Natural History, Chicago (FMNH). We used the taxonomy in Wilson \& Reeder's Mammal Species of the World \citeyearpar{Wilson2005}, except for the recently discovered tenrec species \textit{Microgale jobihely} \citep{Goodman2006}. We photographed all of the tenrec and golden mole skulls available in the collections. This included 31 of the 34 species in the tenrec family \citep{Olson2013} and 12 of the 21 species of golden moles \citep{Wilson2005}.

	We took pictures of the skulls using photographic copy stands. To take possible light variability into account, we took a photograph of a white sheet of paper each day and used the custom white balance function on the camera to set the image as the baseline "white" measurement for those particular light conditions. We photographed the specimens with a Canon EOS 650D camera fitted with a EF 100 mm f/2.8 Macro USM lens and using a remote control (H\"ahnel Combi TF) to avoid camera shake. We photographed the specimens on a black material background with a light source in the top left-hand corner of the photograph. We used small bean bags to hold the specimens in position to ensure that they lay in a flat plane relative to the camera, and used the grid-line function on the live-view display screen of the camera to position the specimens in the centre of each image.

	All skulls were photographed in three views: dorsal, ventral and lateral (right side) (Figure 1). When the right sides of the skulls were damaged or incomplete, we photographed the left sides and later reflected the images \citep[e.g.][]{Barrow2008}. Some specimens were too damaged to use in particular views so there were a different total number of images for each analysis. Our final data sets included photographs of 182 skulls in dorsal view (148 tenrecs and 34 golden moles), 173 skulls in ventral view (141 tenrecs and 32 golden moles) and 171 skulls in lateral view (140 tenrecs and 31 golden moles). After taking the photographs we converted the raw files to binary (grey scale) images and re-saved them as TIFF files \citep[uncompressed files preserve greater detail, ][]{RHOI2013}. Photographs of the specimens from the American Museum of Natural History and the Smithsonian Institute Natural History Museum are available on figshare (dorsal; \citet{Finlay2013d}, ventral; \citet{Finlay2013v} and lateral; \citet{Finlay2013l}). Copyright restrictions from the other museums prevent public sharing of their images but they are available on request.

@@

	When using semilandmark approaches there is a potential problem of over-sampling: simpler structures will require fewer semilandmarks to accurately represent their shape \citep{MacLeod2012}. To ensure that we applied a uniform standard of shape representation to each outline segment (i.e. that simple structures would not be over-represented and more complex features would not be under-represented), we followed the method outlined by MacLeod \citeyearpar{MacLeod2012} to determine the minimum number of semilandmark points which would give accurate representations of morphological shape.

@@

	We grouped the PC scores for tenrecs and golden moles separately so that we could estimate the diversity of each family and then compare the two groups (Figure 1). We compared morphological diversity in two ways. First, we used non parametric multivariate analysis of variance \citep[npMANOVA;][]{Anderson2001} to test whether tenrecs and golden moles occupied significantly different positions within the morphospaces defined by the PC axes that accounted for 95\% of the overall variation in the data \citep[e.g.][]{Serb2011, Ruta2013}. A significant difference between the two Families would indicate that they have unique morphologies which do not overlap. Second, we compared morphological diversity within tenrecs to the diversity within golden moles. We define morphological diversity as the mean Euclidean distance (sum of squared differences) between each species and its family centroid (Figure 3). This is summarised in the equation below where \textit{n} is the number of species in the family, \textit{i} is the number of PC axes and \textit{c} is the average PC score for each axis (the centroid).

@@

	In addition to the differences among the three skull views, our results indicate that, in skulls at least, the overall morphological diversity within tenrecs is not as large as is often assumed \citep[e.g.][]{Eisenberg1969, Olson2013}. Studies of morphological variation are sensitive to the sampling used. If a particular morphotype is over-represented then the similarities among those species will reduce the overall morphological variation within the group \citep{Foote1991}. This appears to be the case for our data; it was only when we included a sub-sample of \textit{Microgale} tenrecs that we found higher morphological diversity in tenrecs compared to golden moles across all three skull analyses (Table 1).
	While there are clear physical differences among family members \citep{Olson2013, Eisenberg1969}, the majority of tenrecs are very morphologically similar \citep{Jenkins2003} so morphological diversity in the family as a whole is not as large as it first appears.

	The goal of our study was to quantify morphological variation in tenrecs instead of relying on subjective assessments of their high morphological diversity \citep{Olson2013, Soarimalala2011, Eisenberg1969}. However, it is difficult to quantify overall morphological diversity because any study is inevitably constrained by its choice of specific traits \citep{Roy1997}. While the skull is widely regarded as a good model for studying morphological variation \citep[e.g.][]{Blagojevic2011, Flores2010, Giannini2010}, quantifying variation in other morphological traits could yield different patterns. Therefore future work should extend our approach beyond skulls to gain a more complete understanding of the overall morphological diversity of tenrecs and golden moles.
	While recognising these limitations, our results provide valuable insights into the differences between subjective and quantitative assessments of morphological diversity.

\section{Conclusions}
	We have presented the first quantitative investigation of morphological diversity in tenrecs. Our results indicate that, overall, tenrec skulls are not more morphologically diverse than golden moles and that similarities among the species rich \textit{Microgale} tenrecs mask signals of higher morphological diversity among the rest of the family. Of course the results presented here are restricted to just one axis of morphological variation and further analysis of other traits is required. However, our findings provide a foundation for future investigations and represent a significant step towards a more quantitative understanding of patterns of morphological and evolutionary diversity in tenrecs.

@@

% Figure 3: Calculating diversity as mean Euclidean distance to family centroid.
% Estimating morphological diversity as the mean Euclidean distance between each species and the family centroid. Every species had scores on the principal components (PC) axes that accounted for 95\% of the variation in the principal components analysis. The number of axes (PCn) varied for each analysis but they were the same within a single analysis. PC scores were used to calculate the Euclidean distance from each species to the family centroid (average PC scores for the entire family). Morphological diversity of the family is the average value of these Euclidean distances.

>>>>

	Unfortunately, morphological diversity is difficult to quantify. Many studies are constrained to measuring the diversity of specific traits rather than overall morphologies \citep{Roy1997}. In addition, our perception of morphological diversity is influenced by the trait being measured, and results may depend on the particular trait being analysed \citep{Foth2012}. Furthermore, linear measurements of morphological traits can restrict our understanding of overall morphological variation; a distance matrix of measurements among specific points is unlikely to give a complete representation of a three dimensional structure \citep{Rohlf1993}. Geometric morphometric approaches can circumvent some of these issues by using a system of Cartesian landmark coordinates to define anatomical points \citep[][and references therein]{Adams2004}. This method captures more of the true, overall anatomical shape of particular structures \citep{Mitteroecker2009}. In particular, two-dimensional geometric morphometric approaches are commonly used to analyse 3D morphological shape and are appropriate for cross-species comparisons \citep[e.g.][]{Muschick2012, Panchetti2008, Wroe2007}. Any bias from 2D representation of a 3D structure is unlikely to be a significant issue for interspecific studies as the overall shape variation among species is geater than discrepancies introduced by using 2D morphometric techniques \citep{Cardini2014}. These more detailed approaches are useful tools for studying patterns of morphological diversity.

	Here we apply geometric morphometric techniques to quantify morphological diversity in a family of small mammals, the tenrecs. Tenrecs (Afrosoricida, Tenrecidae) are a morphologically diverse group that researchers often identify as an example of both convergent evolution and an adaptive radiation \citep{Soarimalala2011, Eisenberg1969}. The family is comprised of 34 species, 31 of which are endemic to Madagascar \citep{Olson2013}. Body masses of tenrecs span three orders of magnitude (2.5 to $>$ 2,000g); a greater range than all other Families, and most Orders, of living mammals \citep{Olson2003}. Within this vast size range there are tenrecs which resemble shrews (\textit{Microgale} tenrecs), moles (\textit{Oryzorictes} tenrecs) and hedgehogs \citep[\textit{Echinops} and \textit{Setifer} tenrecs,][]{Eisenberg1969}. The similarities among tenrecs and other small mammal species include examples of morphological, behavioural and ecological convergence \citep{Soarimalala2011}. Tenrecs are one of only four endemic mammalian clades in Madagascar and the small mammal species they resemble are absent from the island \citep{Garbutt1999}. Therefore, it appears that tenrecs represent an adaptive radiation of species which filled otherwise vacant ecological niches \citep{Soarimalala2011}.

	The claims that tenrecs are an example of both an adaptive radiation and convergent evolution have not been investigated quantitatively. There are qualitative similarities among the hind limb morphologies of tenrecs and several other unrelated species with similar locomotory styles \citep{Salton2009} but the degree of morphological similarity has not been established. Morphological diversity is an important feature of adaptive radiations \citep{Losos2010a} and it also informs our understanding of convergent phenotypes \citep{Muschick2012}. Therefore, it is important to quantify patterns of morphological diversity in tenrecs to gain an insight into their evolution.

	We present the first quantitative study of patterns of morphological diversity in tenrecs. We use geometric morphometric techniques \citep{Rohlf1993} to compare cranial morphological diversity in tenrecs to that of their closest relatives, the golden moles (Afrosoricida, Chrysochloridae). 
	We expect tenrecs to be more morphologically diverse than golden moles because tenrecs occupy a wider variety of ecological niches. The tenrec family includes terrestrial, semi-fossorial, semi-aquatic and semi-arboreal species \citep{Soarimalala2011}. In contrast, all golden moles occupy very similar, fossorial ecological niches \citep{Bronner1995}. Greater ecological variety is often \citep[though not always:][]{McGee2013, Losos2010a} correlated with higher morphological diversity. However, our results reveal that, in skulls at least, morphological diversity in tenrecs is not as great as it first appears.

@@

\subsection{Data collection}
	One of us (SF) collected data from five museums: Natural History Museum, London (BMNH), Smithsonian Institute Natural History Museum, Washington D.C. (SI), American Museum of Natural History, New York (AMNH), Museum of Comparative Zoology, Cambridge M.A. (MCZ) and Field Museum of Natural History, Chicago (FMNH). We used the taxonomy in Wilson \& Reeder's Mammal Species of the World \citeyearpar{Wilson2005}, except for the recently discovered tenrec species \textit{Microgale jobihely} \citep{Goodman2006}. We photographed all of the intact tenrec and golden mole skulls available in the collections. This included 31 of the 34 species in the tenrec family \citep{Olson2013} and 12 of the 21 species of golden moles \citep{Wilson2005}.

	We took pictures of the skulls using photographic copy stands. To take possible light variability into account, we took a photograph of a white sheet of paper each day and used the custom white balance function on the camera to set the image as the baseline "white" measurement for those particular light conditions. We photographed the specimens with a Canon EOS 650D camera fitted with a EF 100 mm f/2.8 Macro USM lens and using a remote control (H\"ahnel Combi TF) to avoid camera shake. We photographed the specimens on a black material background with a light source in the top left-hand corner of the photograph and a scale bar placed below the specimen. We used small bean bags to hold the specimens in position to ensure that they lay in a flat plane relative to the camera, and used the grid-line function on the live-view display screen of the camera to position the specimens in the centre of each image.

	All skulls were photographed in three views: dorsal, ventral and lateral (right side) (Figure 1). When the right sides of the skulls were damaged or incomplete, we photographed the left sides and later reflected the images \citep[e.g.][]{Barrow2008}. Some specimens were too damaged to use in particular views so there were a different total number of images for each analysis. Our final data sets included photographs of 182 skulls in dorsal view (148 tenrecs and 34 golden moles), 173 skulls in ventral view (141 tenrecs and 32 golden moles) and 171 skulls in lateral view (140 tenrecs and 31 golden moles). After taking the photographs we used the Canon Digital Photo Professional software \citep{DPP2013} to convert the raw files to binary (grey scale) images and re-save them as TIFF files \citep[uncompressed files preserve greater detail, ][]{RHOI2013}. Photographs of the specimens from the American Museum of Natural History and the Smithsonian Institute Natural History Museum are available on figshare (dorsal; \citet{Finlay2013d}, ventral; \citet{Finlay2013v} and lateral; \citet{Finlay2013l}). Copyright restrictions from the other museums prevent public sharing of their images but they are available from the authors on request.

@@

	When using semilandmark approaches there is a potential problem of over-sampling: simpler structures will require fewer semilandmarks to accurately represent their shape \citep{MacLeod2012}. To ensure that we applied a uniform standard of shape representation to each outline segment (i.e. that simple structures would not be over-represented and more complex features would not be under-represented), we followed the method outlined by MacLeod \citeyearpar{MacLeod2012} to determine the minimum number of semilandmark points which would give accurate representations of morphological shape. We used 54 points for skulls in dorsal view (10 landmarks, 44 semilandmarks across 4 curves), 73 points for skulls in ventral view (13 landmarks, 60 semilandmarks) and 44 points for skulls in lateral view (9 landmarks and 35 semilandmarks across 2 curves) (see Figure 2 and the supplementary material for more details).

@@

	We grouped the PC scores for tenrecs and golden moles separately so that we could estimate the diversity of each family and then compare the two groups (Figure 1). We compared morphological diversity in two ways. First, we used non parametric multivariate analysis of variance \citep[npMANOVA;][]{Anderson2001} to test whether tenrecs and golden moles occupied significantly different positions within the morphospaces defined by the PC axes that accounted for 95\% of the overall variation in the data \citep[e.g.][]{Serb2011, Ruta2013}. A significant difference between the two Families would indicate that they have unique morphologies which do not overlap. Second, we compared morphological diversity within tenrecs to the diversity within golden moles.

	Morphological diversity (variation in form) is more commonly referred to as morphological disparity \citep{Foote1997}. There are many different methods for measuring disparity. Calculations based on summary (principal component) axes of shape variation are popular \citep[e.g.][]{Ruta2013, Foth2012, Brusatte2008, Wainwright2007} while other methods include calculating disparity directly from Procrustes shape variables \citep{Zelditch2012} or rate-based approaches which depend on phylogenetic branching patterns \citep[e.g.][]{Price2013, Price2010, OMeara2006}. There is no single best method of measuring disparity \citep{Ciampaglio2001} and each method makes different assumptions which are appropriate for different situations. Therefore, for clarity, we have chosen to measure variation in physical form using a clear, easily-interpretable method which captures variation in morphological diversity.

	We define morphological diversity as the mean Euclidean distance (sum of squared differences) between each species and its family centroid (Figure 3). This is summarised in the equation below where \textit{n} is the number of species in the family, \textit{i} is the number of PC axes and \textit{c} is the average PC score for each axis (the centroid).

@@

	In addition to the differences among the three skull views, our results indicate that, in skulls at least, the overall morphological diversity within tenrecs is not as large as is often assumed \citep[e.g.][]{Eisenberg1969, Olson2013}. Studies of morphological variation are sensitive to the sampling used. If a particular morphotype is over-represented then the similarities among those species will reduce the overall morphological variation within the group \citep{Foote1991}. This appears to be the case for our data; it was only when we included a sub-sample of \textit{Microgale} tenrecs that we found higher morphological diversity in tenrecs compared to golden moles across all three skull analyses (Table 1).
	While there are clear physical differences among family members \citep{Olson2013, Eisenberg1969}, the majority of tenrecs (the \textit{Microgale}) are very morphologically similar \citep{Jenkins2003} so morphological diversity in the family as a whole is not as large as it first appears.

	The goal of our study was to quantify morphological variation in tenrecs instead of relying on subjective assessments of their high morphological diversity. However, it is difficult to quantify overall morphological diversity because any study is inevitably constrained by its choice of specific traits \citep{Roy1997}. While the skull is widely regarded as a good model for studying morphological variation \citep[e.g.][]{Blagojevic2011, Flores2010, Giannini2010}, quantifying variation in other morphological traits could yield different patterns. Therefore future work should extend our approach beyond skulls to gain a more complete understanding of the overall morphological diversity of tenrecs and golden moles.
	While recognising these limitations, our results provide valuable insights into the differences between subjective and quantitative assessments of morphological diversity.

\section{Conclusions}
	We have presented the first quantitative investigation of morphological diversity in tenrecs. Our results indicate that, overall, tenrec skulls are not more morphologically diverse than golden moles and that similarities among the species rich \textit{Microgale} tenrecs mask signals of higher morphological diversity among the rest of the family. Of course the results presented here are restricted to just one aspect of morphological variation and further analysis of other traits is required. However, our findings provide a foundation for future investigations and represent a significant step towards a more quantitative understanding of patterns of morphological and evolutionary diversity in tenrecs.

@@

% Figure 3: Calculating diversity as mean Euclidean distance to family centroid.
% Estimating morphological diversity as the mean Euclidean distance between each species and the family centroid. Every species had scores on the principal components (PC) axes that accounted for 95% of the variation in the principal components analysis. The number of axes (PCn) varied for each analysis but they were the same within a single analysis. PC scores were used to calculate the Euclidean distance from each species to the family centroid (average PC scores for the entire family). Morphological diversity of the family is the average value of these Euclidean distances.
